Scheduling and Amortization for MPC

### 2021-11-21 paper_SIMD.tex
Basic vectorization, cleaner version.

% *** SPECIALIZED LIST PACKAGES ***
\usepackage{algorithmic}

@@

%
% The "author" command and its associated commands are used to define the authors and their affiliations.
% Of note is the shared affiliation of the first two authors, and the "authornote" and "authornotemark" commands
% used to denote shared contribution to the research.
\author{Muhammad Ishaq}
\authornote{This work was done in part while the author was at RPI.}
\email{m.ishaq@ed.ac.uk}
%\orcid{1234-5678-9012}
%\author{G.K.M. Tobin}
%\authornotemark[1]
%\email{webmaster@marysville-ohio.com}
\affiliation{%
    \institution{University of Edinburgh}
    %\streetaddress{P.O. Box 1212}
    \city{Edinburgh}
    \state{Scotland}
%    \postcode{43017-6221}
}

@@

@@

In this short paper we define the problem of \emph{optimal parallelization} for Multi-party Computation (MPC), 
set it apart from classical work on parallelization for High-performance Computing (HPC), present preliminary 
results, and set directions for future research.

\ana{1. Intro to MPC (from a PL point of view). Define the problem.}

\ana{2. Difference with HPC. Why "optimal HPC" = "minimal delay" is NOT "optimal MPC"}

\ana{3. Results statement.}

\ana{4. Future directions.}

\ana{Scope of paper: Should we expand towards 1) nested loops, 2) conversions, 3) implementation (what will we be saying with an implementation?)}

% no \IEEEPARstart

@@

\paragraph{Problem 1:} Given a dependence graph of MPC instructions, find the \emph{loop-body} schedule 
with \emph{minimal cost}.

%\begin{figure*}[tbhp]
\begin{figure*}[tbhp]
\begin{tabular}{llll}
\begin{minipage}[b]{4.25cm}

\begin{minted}[fontsize=\footnotesize, linenos, numbersep=5pt, escapeinside=||]{java}
for (int i=1; i<=N; i++) {
  A: A[i] = f_1(B[i]);
  B: B[i] = f_2(A[i],D[i-1]);
  C: C[i] = f_3(A[i],D[i-1]);
  D: D[i] = f_4(B[i],C[i]);
}
\end{minted}
\end{minipage}

&

\begin{minipage}[b]{4.25cm}
\includegraphics[width=0.5\textwidth]{figs/dependenceGraph.pdf}
\end{minipage}

&

\begin{minipage}[b]{4.25cm}
\includegraphics[height=0.15\textheight]{figs/MPCloop.pdf}
\end{minipage}

&

\begin{minipage}[b]{4.25cm}
\includegraphics[height=0.20\textheight]{figs/MPCaccrossLoop.pdf}
\end{minipage}
\\

(a) Source 
&
(b) Dependence graph
&
(c) Loop-body schedule
&
(d) Across-loop schedule

\end{tabular}
\caption{(a) shows the source code, taken from Aiken and Nicolau~\cite{Aiken:1988}, 
(b) shows the dependence graph. The solid edges denote within-iteration ordering constraints, e.g., A must run before B
because B depends on A. The dashed edges denote across-iteration ordering constraints, e.g., B(2), i.e., B in iteration 2
depends on D(1), i.e., D in iteration 1; in the HPC literature these are known as loop-carried dependencies. 
(c) shows the optimal loop-body schedule, and (d) shows the optimal across-loop schedule. A(1), B(1), etc. denote 
instruction A in iteration 1, etc.}
\label{fig:MPCexample}\vspace{-2ex}
\end{figure*}

\paragraph{Problem 2:} Given a dependence graph of MPC instructions representing a loop body (or given a loop-body schedule), 
find the \emph{across-loop} schedule with  \emph{minimal cost}.

\figref{fig:MPCexample} illustrates the problem with a code example from the classical high-performance computing (HPC) literature. 
\figref{fig:MPCexample}(c) shows the optimal loop-body schedule (answer to Problem 1). Instructions B and C can be scheduled in parallel, 
however, the schedule does not necessarily benefit form parallelization/amortization. If B and C (functions $f_1$ and $f_2$) are the \emph{same}
MPC instruction (e.g., both are MUX, MUL, ADD, or CMP, etc.), then they \emph{can} benefit from parallelization/amortization. Given
our assumptions stated above, the cost of running the loop-body schedule is 3. However, if B and C are different MPC-instructions,
e.g., B is MUL, but C is MUX, then the cost of running the schedule is 4.

\figref{fig:MPCexample}(d) shows the optimal across-loop schedule (answer to Problem 2). We can schedule all A-instructions, across all iterations, 
in parallel, as A depends only on the initial value of B. We cannot do better but schedule B, C, and D sequentially, as they each depend on values
computed in the previous iteration. Again, if B and C are the same instruction, the optimal across-loop schedule will have cost of $1 + 2N$, 
and if B and C are NOT the same instruction, then it will have cost $1+3N$.

\ana{Mention difficulties. We start from IMP-like general-purpose PL syntax with array reads and writes. We translate the program into a dependence 
graph of MPC instructions. Then we proceed to compute the schedule.}

@@

In this section, we state the classical HPC scheduling problem, and argue the MPC scheduling problem poses unique challenges.

Similar problems arise in the HPC setting.

%\begin{figure*}[tbhp]
\begin{figure*}[tbhp]
\begin{tabular}{lll}

\begin{minipage}[b]{4.25cm}
\includegraphics[width=0.4\textwidth]{figs/HPCloop.pdf}
\end{minipage}

&

\begin{minipage}[b]{4.25cm}
\includegraphics[width=0.4\textwidth]{figs/HPCdoacrossLoop.pdf}
\end{minipage}

&

\begin{minipage}[b]{4.25cm}
\includegraphics[width=0.6\textwidth]{figs/HPCgreedyLoop.pdf}
\end{minipage}

\\

(a) Loop-body schedule
&
(b) Doacross across-loop schedule
& 
(c) Greedy across-loop schedule

\end{tabular}
\caption{Optimal HPC schedules.}
\label{fig:HPCexample}\vspace{-2ex}
\end{figure*}

\paragraph{Problem 1:} Given a loop body (i.e., dependence graph), find the loop-body schedule with \emph{minimal delay}.

\paragraph{Problem 2:} Given a loop body (or a loop-body schedule), find the \emph{doacross} schedule with minimal delay.

\paragraph{Problem 3:} Given a loop body (or a loop-body schedule), find the \emph{greedy schedule} with minimal delay.

\figref{fig:HPCexample} illustrates the problem in HPC. \figref{fig:HPCexample}(a) shows the optimal loop-body schedule
which results in cost of 3. Again, we assume that all instructions have the same cost, a unit of 1. The key difference with the 
MPC schedule is that regardless of whether B and C are the same instruction opcode or not, B and C can be scheduled 
to run in parallel and the overall running time benefits from parallelization.

The HPC literature has considered different kinds of across-loop scheduling. One scheduling, called \emph{doacross}~\cite{aiken,cytron},
requires that all instructions in an iteration are scheduled on the same processor. The optimal doacross schedule for our example
is shown in \figref{fig:HPCexample}(b) (taken directly from~\cite{aiken}). Aiken and Nicolau proposed a new schedule,
which they called \emph{greedy} schedule, where instructions in an iteration can be scheduled on different processors. 
The optimal greedy schedule for our example, which achieves better minimal delay than the optimal doaccross schedule, 
is shown in \figref{fig:HPCexample}(c) (again, taken directly from~\cite{aiken}).

\ana{Here we have to make the argument that MPC is different from HPC.}
The example illustrates that optimal MPC scheduling and optimal HPC scheduling are different problems.
To summarize, scheduling and amortization in MPC is a different problem due to two key differences: (1) there 
is unlimited bandwidth while in HPC, we typically have a limited number of processors we can use to 
schedule instructions in parallel, and (2) we benefit from amortization only if the set of instructions scheduled
in parallel are the same; in contrast, in HPC any two instructions can be scheduled in parallel.

\ana{Have to make the argument that in MPC all stmts are executed. Thus, the vectorization problem is
immediately simpler!}

%\section{\bf Preliminaries}
%\label{sec:preliminaries}

\ana{Relevant papers/books:

\begin{enumerate}

\item Book: "Scheduling and automatic parallelization", Springer 2000, by Alain Darte, Yves Robert and Frederic Vivien. 
This book describes scheduling and parallelization in HPC: given a dependence graph and unlimited 
number of processors, one can compute an optimal (minimal delay) schedule in polynomial time. 
Given a finite number of processors, computing an optimal schedule becomes NP-hard.

\item https://theory.stanford.edu/~aiken/publications/papers/pldi88.pdf. "Optimal loop parallelization", PLDI 1988, by
Alex Aiken and Alex Nicolau. 
Given a dependence graph of the loop body, computes an optimal parallelization across loop iterations. 
The optimality metric is minimal delay. E.g., assuming 2 iterations, the delay is measured by the number 
of instructions in iteration 2 that are lagging behind iteration 1.

\item https://www.cs.indiana.edu/~achauhan/Teaching/B629/2006-Fall/CourseMaterial/1998-popl-knobe-arrayssa.pdf.
"Array SSA", POPL 1998, by Kathleen Knobe and Vivek Sarkar. This paper defines an SSA form for arrays which I think 
is crucial for parallelization across loop iterations.

\end{enumerate}
}

\subsection{Precise Problem Statement}
\ana{TODO}

\section{Loop Body Parallelization}
\label{sec:parallelization_within_loop}
%\input{../sections/analysis.tex}

We first consider the answer to Problem 1 (in MPC). 
We argue that optimal parallelization within a loop body is NP-complete (under the above stated assumptions). 
Therefore, compilers must resort to heuristics to compute a schedule for the instructions within a loop body.

We consider two operations, call them $A$ and $M$. Each instruction in the program is either an $A$-instruction or
an $M$-instruction. In order to benefit from parallelization/amortization, we must schedule two or more 
$A$-instructions in the same parallel node (or two or more $M$-instructions in the same parallel node). 
Scheduling $A$-instructions in parallel with $M$-instruction does not benefit from amortization.
It incurs the exact same cost as scheduling the $A$-instructions in a node $P_A$, scheduling the $M$-instructions 
in a node $P_M$, and having $P_A$ precede $P_M$ in the parallel schedule. This is the difference 
between classical scheduling, as studied in parallel computing, and MPC scheduling.

We make the following assumptions:

\begin{enumerate}

\item $A$ and $M$ are of equal cost, 1 unit.
\item There is unlimited bandwidth---i.e., a single $A$-instrution (or $M$-instruction) costs as much as $N$ amortized $A$-instructions 
(or $M$-instructions), namely 1 unit.

\end{enumerate}

Consider a loop body that consists of $n$ sequences: $S_1$, ... $S_n$ of $A$ and $M$ instructions. 
More precisely, the loop body is such that its instructions can be grouped into such sequences. 
$S_1$, ... $S_n$ can execute in parallel, however, all instructions within a sequence must 
execute sequentially. For example, consider the three sequences (the right arrow indicates a \emph{dependence},
meaning that the source node must execute before the target node): 
\begin{enumerate}
\item $A \rightarrow M \rightarrow A$
\item $A \rightarrow A \rightarrow A$
\item $M \rightarrow A \rightarrow M$
\end{enumerate}

A \emph{schedule} $P: P_1 \rightarrow P_2 \dots \rightarrow P_k$ is such that for each sequence 
$S_i$ in the set, if $S_i[k]$ precedes $S_i[k']$ in $S_i$ then $S_i[k]$ is scheduled in node $P_l$, $S_i[k]$ 
is scheduled in node $P_{l'}$, and $P_l$ precedes $P_{l'}$ in $P$.

The cost of a schedule $P$ is 
\[\mathit{cost}(P) = \sum_{i=1}^k \mathit{cost}(P_i)\]
where $\mathit{cost}(P_i) = 1$ if $P_i$ consists of $A$-instructions only, or $M$-instructions only, 
and $\mathit{cost}(P_i) = 2$ if $P_i$ mixes $A$-instructions and $M$-instructions.

The problem is to find a schedule $P$ with \emph{minimal cost}. For example, 
a schedule with minimal cost for the sequences above is
\[ A(1), A(2) \rightarrow M(1), A(2), M(3) \rightarrow A(1), A(2), A(3) \rightarrow M(3) \]
(The parentheses above indicate the sequence where the instruction comes from: (1), (2), or (3).)
The cost of this schedule is 5.

The problem of finding a schedule with minimal cost 
easily reduces to the problem of finding a \emph{shortest common supersequence}, 
a known NP-complete problem \ana{citation needed}. To see the reduction, 
suppose $P$ is a schedule with minimal cost (computed by a black-box algorithm). 
We can derive a schedule $P'$ with the same cost as $P$, by mapping each mixed node $P_i \in P$ 
to two consecutive nodes in $P'$: an $A$-instruction node followed by an $M$-instruction node.
Clearly, $P'$, which now is a sequence of $A$'s and $M$'s, is a supersequence of each sequence 
$S_i$, i.e., $P'$ is a common supersequence 
of $S_1 \dots S_n$. It is also a shortest common supersequence. (To see this, suppose, a 
shorter common supersequence, $P''$, exists. $P''$ is a schedule of $S_1 \dots S_n$
and the $\mathit{cost}(P'')$ equals the length of $P''$. Since $P'$ is longer than
$P''$, and $\mathit{cost}(P') = \mathit{cost}(P)$, that means $\mathit{cost}(P'') < \mathit{cost}(P)$, 
which is a contradiction since $P$ is a schedule with minimal cost.)

\ana{Question: are these two assumptions too strong?}

\ana{Assumption 1 can be relaxed --- e.g., if $A$ is 5 times cheaper than $M$, 
we can construct a program where all $A$'s are sequences of five consecutive 
$A$'s, which reduces to the previous problem.}

\ana{Assumption 2, I am not sure if this is acceptable, or maybe it is too strong. 
My intuition is that incorporating into the cost a dependence on the number 
of instructions in each parallel node, is going to make the problem harder, not easier...}

\section{Across-Loop Parallelization} 
\label{sec:optimal_loop_parallelization}
%\input{../sections/optimal_loop_parallelization.tex}

We now consider the answer to Problem 2 (in MPC).

\begin{comment}
\subsection{Overview and Problem Statement}

The problem is, given a loop-body schedule, compute the across-loop schedule
with minimal cost. The following examples illustrate. The first example is the inner 
product computation, and the second is Aiken and Nicolau's example.

\begin{center}
\begin{lstlisting}[language=Java]
A = ...
B = ...
ip = 0;
for (i=0; i<N; i++) {
  if (i%d(5)==0) para{t[i]=A[i]*B[i],...,t[i+d(5)-1]=A[i+d(5)-1]*B[i+d(5)-1]}
  if (i%d(6)==0) para{ip = ip + t[i]};
}  
\end{lstlisting}
\end{center}

The above notation is a convenient way to specify a schedule. Each statement $s$ in the loop body sequence 
is parallelized at a rate $d(s)$ (we assume that the loop body is a straight-line sequence which is the case in MPC). 
Notation  
\[{\sf if \; (i\%d(s)\!==\!0) \; para\{s(i),s(i+1),...,s(i+d(s)-1)\}}\]
means that the first $d(s)$ iterations of statement $s$, denoted by $s(i),...s(i+d(s)-1)$, are executed in a parallel block; 
then the following $d(s)$ iterations are executed in a parallel block, and so on. The execution of $d(s)$ statements $s$ in 
a block benefits from amortization of the operation at $s$.
If {\sf i\%d(s)==0} evaluates to false, then the $i$-th iteration skips $s$, because $s(i)$ has executed earlier in a parallel block.

In the above example {\sf d(5)} is N, meaning that all N multiplications are executed in parallel.
In contrast, {\sf d(6)} is 1, meaning that the summation statements are executed sequentially.

A more interesting example involves array reads and writes. We return to the Aiken and Nicolau
example:

\begin{lstlisting}[language=Java]
for (i=0; i<N; i++) {
  if (i%N==0) para{A[i]=f1(B[i]),...,A[i+N-1]=f1(B[i+N-1])};
  if (i%1==0) para{B[i]=f2(A[i],D[i-1])};
  if (i%1==0) para{C[i]=f3(A[i],D[i-1])};
  if (i%1==0) para{D[i]=f4(B[i],C[i])};
}	
\end{lstlisting}

The above across-loop schedule entails that the first instruction is parallelized---all
N instructions A can be computed at once because each depends only on {\sf B[i]}, 
which at the point {\sf A[i]} is computed, refers to a value assigned outside of the loop. 
In contrast, subsequent instructions are executed sequentially. Take B, for example. 
There is a dependence ``cycle'' from {\sf B[i]} to {\sf D[i]} then {\sf B[i+1]} in the subsequent 
iteration. (The cycle is shown in~\figref{fig:MPCexample}(b).) {\sf B[i]}'s cannot be computed 
in parallel because each one depends on the {\sf B[i]} value computed in the previous iteration.

Now consider the above code but slightly modified. {\sf B[i]} and {\sf C[i]} are now 
computed based on the value of {\sf D[i-2]} assigned two iterations earlier.

\begin{tabular}{lll}
\begin{lstlisting}[language=Java]
for (i=1; i<=N; i++) {
  A[i]=f1(B[i]);
  B[i]=f2(A[i],D[i-2]);
  C[i]=f3(A[i],D[i-2]);
  D[i]=f4(B[i],C[i]);
}

\end{lstlisting}
& 
~
&
\begin{lstlisting}[language=Java]
for (i=0; i<N; i++) {
  if (i%N==1) para{A[i]=f1(B[i]),...};
  if (i%2==1) para{B[i]=f2(A[i],D[i-2]),...};
  if (i%2==1) para{C[i]=f3(A[i],D[i-2]),...};
  if (i%2==1) para{D[i]=f4(B[i],C[i]),...};
}	
\end{lstlisting}

\end{tabular}

{\sf B[i]} still depends on a value computed in earlier iterations, i.e., there are loop-carried dependences, 
but the value, D[i-2] is two iterations earlier. Thus, we can take parallelization blocks of 2, e.g., computing 
{\sf B[0],B[1]} in a parallel block, then {\sf B[2],B[3]}, etc.
Note however, that we cannot take a parallelization block larger than 2.
%The value of B will depend on a value being computed in parallel.

The key problem that we address in this paper is computing the maximal rate $d(s)$ allowed by program dependences 
for each statement $s$. Program dependences give rise to a partitioning of the sequence of {\sf para} blocks into a DAG of
strongly connected components. (We elaborate on this and define the partitioning precisely later in the paper.)
In the above example, there are 2 SCCs---one component consists of just the {\sf para} block that computes 
{\sf f1(B[i])} and the other component consists of the remaining three {\sf para} blocks. The three para blocks
are interdependent and hence, they are in the same component.

It is an invariant of our approach that the rates of all blocks in a SCC are the same, $d(s_1) = d(s_2) = d(\mathit{SCC})$
for each $s_1, s_2 \in \mathit{SCC}$. Unfortunately, as we argued earlier scheduling within a SCC is NP-complete, 
and also, scheduling of the components in the DAG is NP-complete. Assuming that we have an optimal schedule 
for each SCC, the sum of the costs of all SCCs is an upper bound on the optimal cost and the sum of the longest chain
in the DAG is a lower bound on the optimal cost. In the special case when the DAG is a straight line, which is the case 
is the typical case \ana{how many of the actual benchmarks???} the sum the costs of all SCCs is the \emph{minimal} cost.

\ana{Because the results are somewhat negative, i.e., thing is NP-complete, we'd need some experiments, at least 
analysis of loop structure in different benchmarks. I am pretty sure the structure is straight-line DAG with very simple SCCs, 
hence, we do have an optimal schedule anyway.}

\subsection{Preliminaries}

%\begin{figure*}[tbhp]
\begin{figure}[tbhp]
\begin{tabular}{ll}
\begin{minipage}[b]{4.25cm}

>>>>

% *** SPECIALIZED LIST PACKAGES ***
%\usepackage{algorithm2e}
\usepackage{algorithmic}

@@

%
% The "author" command and its associated commands are used to define the authors and their affiliations.
% Of note is the shared affiliation of the first two authors, and the "authornote" and "authornotemark" commands
% used to denote shared contribution to the research.

\author{Benjamin Levy}
\email{levyb3@rpi.edu}
\affiliation{%
    \institution{Rensselaer Polytechnic Institute}
    %\streetaddress{1 Th{\o}rv{\"a}ld Circle}
    \city{Troy}
    \state{New York}
    %\country{Iceland}
}

\author{Benjamin Sherman}
\email{shermb@rpi.edu}
\affiliation{%
    \institution{Rensselaer Polytechnic Institute}
    %\streetaddress{1 Th{\o}rv{\"a}ld Circle}
    \city{Troy}
    \state{New York}
    %\country{Iceland}
}

@@

\author{Muhammad Ishaq}
\authornote{This work was done in part while the author was at RPI.}
\email{m.ishaq@ed.ac.uk}
%\orcid{1234-5678-9012}
%\author{G.K.M. Tobin}
%\authornotemark[1]
%\email{webmaster@marysville-ohio.com}
\affiliation{%
    \institution{University of Edinburgh}
    %\streetaddress{P.O. Box 1212}
    \city{Edinburgh}
    \state{Scotland}
%    \postcode{43017-6221}
}

@@

@@

@@

\section{\bf Preliminaries}
\label{sec:preliminaries}

We assume arbitrarily nested loops in the MPC-source IR and read-only arrays. We assume that loops range from 0 to some constant N. Arrays are linearized (row-major order as in MOTION) and accesses are via functions of the induction variables of the enclosing loops. We write $i_1,i_2,...i_k$ to denote the loop nest: $i_1$ is the outermost loop, $i_2$, is immediately nested in $i_1$, and so on until $i_k$. As an example, a statement nested in $i_1,i_2,...i_k$ can access array $A[f(i_1,i_2,...i_k)]$. We write $A[i_1,i_2,...i_k]$ interchangeably.

The dependence graph computes def-use edges:

\begin{itemize}
\item same-level $X\rightarrow Y$ where $X = ...$ and $Y = ...$ are in the same loop nest $i_1,i_2,...i_k$. E.g., $\texttt{d} = \mathit{SUM}(\texttt{S}[i,j], \texttt{C}[j])$ to $\texttt{p} = \mathit{MUL}(\texttt{d},\texttt{d})$ in Biometric is a same-level edge. 
\item outer-to-inner $X\rightarrow Y$ where $X$ is in an outer loop nest, $i_1,i_2,...i_j$, and $Y$ is in an inner one, $i_1,i_2,...i_j,...,i_k$.
\item inner-to-outer $X\rightarrow Y$ where $X$ is in an inner loop nest, $i_1,i_2,...i_j,...,i_k$, and $Y$ is in an outer one, $i_1,i_2,...i_j$. E.g. $\texttt{sum}_1 = \texttt{ADD}(\texttt{sum}_0,\texttt{p})$ to
$\texttt{c} = \mathit{CMP}(\texttt{sum}_1,\texttt{min}_0)$ is an inner-to-outer edge.
\item same-level back-edge $X\rightarrow Y$. $Y$ is a phi-node in the header of the loop and $X$ is a definition of the variable in the loop body. E.g., $\texttt{min}_1 = \mathit{MUX}(\texttt{c},\texttt{sum}_1,\texttt{min}_1)$ to $\texttt{min}_0 = \phi(\texttt{min}_1, 10000)$ in Biometric is a same-level back-edge.
\item outer-to-inner back-edge $X\rightarrow Y$: $X$ and $Y$ are both $\phi$-nodes for some variable. The source $X$ is in an immediately nested loop of $Y$'s loop. 
 \end{itemize}

We also define the $\mathit{closure}(n)$ where $n$ is a phi-node. Intuitively, it computes the set of statements that form a cycle in the loop with $n$. The closure of $n$ is defined as follows:
\begin{itemize}
\item $n$ is in $\mathit{closure}(n)$
\item $X$ is in $\mathit{closure}(n)$ if there is a same-level path from $n$ to $X$, and $X \rightarrow n$ is a same-level back-edge. 
\item $Y$ is in $\mathit{closure}(n)$ if there is a same-level path from $n$ to $Y$ and there is a same-level path from $Y$ to some $X$ in $\mathit{closure}(n)$.
\end{itemize}

\section{Basic Vectorization}

\ana{Add back-edges into Phase 1. A back-edge from a non-phi-node in loop $i$ to a phi-node in loop $i$'s header is a same-level edge. The only difference with the handling of normal def-use is that the operand index will become $i-1$. A back edge from a phi-node in loop $j$ to a phi-node in loop $i$, where $j$ is immediately nested in $i$ is an inner-to-outer edge and will require dropping dimension. We can show these are the only kinds of back-edges that may occur.}

%\subsection{Algorithm (Still work in progress.)}

\begin{algorithmic}

\STATE \COMMENT {Phase 1: Raise dimension of scalar variables to corresponding loop nest. We may assume linear traversal of the MPC-source. \ana{Double check!}}

\FOR {each MPC $\mathit{stmt}: X = \mathit{Op}(Y_1,Y_2)$ in loop $i_1,...,i_k$}
\FOR {each $Y_i$}
\STATE {case def-use edge $Y_i \rightarrow X$ of} \\
\STATE \hspace{0.25cm} {same-level $\rightarrow$ add $Y_i = Y_i$} \\
\STATE \hspace{0.25cm} {outer-to-inner $\rightarrow$ add $Y'_i[i_1,...,i_k] = \mathit{raise\_dim}(Y_i)$} 
\STATE \hspace{0.25cm} {inner-to-outer $\rightarrow$ add $Y'_i[i_1,...,i_k] = \mathit{drop\_dim}(Y_i)$} 
\ENDFOR

\COMMENT {Optimistically vectorize all. $\stackrel{\rightarrow}{i}$ means vectorized dimension.}
\STATE {change to $X[\stackrel{\rightarrow}{i_1},...\stackrel{\rightarrow}{i_k}] = \mathit{Op}(Y'_1[i_1,...,i_k],Y'_2[i_1,...,i_k])$ }
\ENDFOR

\COMMENT {Phase 2: Recreating FOR loops for cycles; vectorizable statements hoisted up.}

\FOR {each dimension $d$ from highest to 0}
\FOR {each $\phi$-node $n$ in loop $i_1,...,i_k$}
\STATE {compute $\mathit{closure}(n)$}
\WHILE {there are closure $cl_1$ and $cl_2$ that intersect}
\STATE {merge $cl_1$ and $cl_2$}
\ENDWHILE 
\FOR {each closure $cl$ (after merge)}
\STATE {create FOR $i_k = 0; ...$ loop}  
\STATE {add $\phi$-nodes in $cl$ to header block} 
\STATE {add statements in $cl$ to loop body in some order of dependences} 
\STATE \COMMENT {Dimension is not vectorizable: }
\STATE {change $\stackrel{\rightarrow}{i_k}$ to $i_k$ in all statements in loop}
\STATE {treat FOR loop as monolith node: change def-use edges accordingly.}
\ENDFOR
\ENDFOR
\ENDFOR

\COMMENT {Phase 3:}
\STATE {add SIMD for simdified dimensions}

\end{algorithmic}

\subsection{Example: Biometric}

\subsubsection{MPC-source}

Below is the MPC-source resulting from Benjamin's analysis. It gives rise to a corresponding dependence graph. (Not shown.)

\begin{algorithmic}

\STATE \COMMENT {Begin of outer loop $i$.}
\STATE {$\texttt{index}_0 = \phi(\texttt{index}_1, -1)$} 
\STATE {$\texttt{min}_0 = \phi(\texttt{min}_1, 10000)$}

\STATE \COMMENT {Begin of inner loop $j$.}
\STATE {$\texttt{sum}_0 = \phi(\texttt{sum}_1, 0)$}
\STATE {$\texttt{d} = \mathit{SUM}(\texttt{S}[i,j], \texttt{C}[j])$}
\STATE {$\texttt{p} = \mathit{MUL}(\texttt{d},\texttt{d})$}
\STATE {$\texttt{sum}_1 = \texttt{ADD}(\texttt{sum}_0,\texttt{p})$}
\STATE \COMMENT {End of inner loop $j$.}
\STATE {$\texttt{c} = \mathit{CMP}(\texttt{sum}_1,\texttt{min}_0)$}
\STATE {$\texttt{index}_1 = \mathit{MUX}(\texttt{c},\texttt{index}_1,i)$}
\STATE {$\texttt{min}_1 = \mathit{MUX}(\texttt{c},\texttt{sum}_1,\texttt{min}_1)$}
\STATE \COMMENT {End of outer loop $i$.}

\end{algorithmic}

\subsubsection{Phase 1 of Vectorization algorithm:}

The transformation preserves the dependence edges. It raises the dimensions of scalars and optimistically vectorizes all operations. The last phase discovers loop-carried dependences and removes affected vectorization.

\begin{algorithmic}

\STATE \COMMENT {Begin of outer loop $i$.}
\STATE {$\texttt{index}_0[i] = \phi(\texttt{index}_1[i-1], -1)$} 
\STATE {$\texttt{min}_0[i] = \phi(\texttt{min}_1[i-1], 10000)$}

\STATE \COMMENT {Begin of inner loop $j$.}
\STATE {$\texttt{sum}_0[i,j] = \phi(\texttt{sum}_1[i,j-1], [0,0,...])$}
\STATE {$\texttt{C}'[i,j] = \mathit{raise\_dim}(\texttt{C}[j],i)$}
\STATE {$\texttt{d}[\stackrel{\rightarrow}{i},\stackrel{\rightarrow}{j}] = \mathit{SUM}(\texttt{S}[\stackrel{\rightarrow}{i},\stackrel{\rightarrow}{j}], \texttt{C}'[\stackrel{\rightarrow}{i},\stackrel{\rightarrow}{j}])$}
\STATE {$\texttt{p}[\stackrel{\rightarrow}{i},\stackrel{\rightarrow}{j}] = \mathit{MUL}(\texttt{d}[\stackrel{\rightarrow}{i},\stackrel{\rightarrow}{j}],\texttt{d}[\stackrel{\rightarrow}{i},\stackrel{\rightarrow}{j}])$}
\STATE {$\texttt{sum}_1[\stackrel{\rightarrow}{i},\stackrel{\rightarrow}{j}] = \mathit{ADD}(\texttt{sum}_0[\stackrel{\rightarrow}{i},\stackrel{\rightarrow}{j}],\texttt{p}[\stackrel{\rightarrow}{i},\stackrel{\rightarrow}{j}])$}
\STATE \COMMENT {End of inner loop $j$.}
\STATE {$\texttt{sum}'_1[i] = \mathit{drop\_dim}(\texttt{sum}_1[i,j],j)$}
\STATE {$\texttt{c}[\stackrel{\rightarrow}{i}] = \mathit{CMP}(\texttt{sum}'_1[\stackrel{\rightarrow}{i}],\texttt{min}_0[\stackrel{\rightarrow}{i}])$}
\STATE {$\texttt{index}_1[\stackrel{\rightarrow}{i}] = \mathit{MUX}(\texttt{c}[\stackrel{\rightarrow}{i}],\texttt{index}_1[\stackrel{\rightarrow}{i}],[0,1,...N])$}
\STATE {$\texttt{min}_1[\stackrel{\rightarrow}{i}] = \mathit{MUX}(\texttt{c}[\stackrel{\rightarrow}{i}],\texttt{sum}'_1[\stackrel{\rightarrow}{i}],\texttt{min}_1[\stackrel{\rightarrow}{i}])$}
\STATE \COMMENT {End of outer loop $i$.}

\end{algorithmic}

\subsubsection{Phase 2}

This phase analyzes statements from the innermost loop to the outermost. The key point is to discover loop-carried dependencies and re-introduce loops whenever dependencies make this necessary.

Starting at the inner phi-node $\texttt{sum}_0[i,j] = \phi(\texttt{sum}_1[i,j], [0,0,...])$, the algorithm first computes it's closure. The closure amounts to the phi-node itself and $\texttt{sum}_1[\stackrel{\rightarrow}{i},\stackrel{\rightarrow}{j}] = \mathit{ADD}(\texttt{sum}_0[\stackrel{\rightarrow}{i},\stackrel{\rightarrow}{j}],\texttt{p}[\stackrel{\rightarrow}{i},\stackrel{\rightarrow}{j}])$, accounting for the loop-carried dependency of the computation of \texttt{sum}. The algorithm replaces this closure with a FOR loop on $j$ removing vectoriaztion on $j$. Note that the SUB and MUL computations remain outside of the loop as they do not depend on phi-nodes that are part of cycles. The algorithm adds same-level edges, one from $\texttt{p}[\stackrel{\rightarrow}{i},\stackrel{\rightarrow}{j}] = ...$ to the monolithic FOR-loop node, and one from the FOR-loop node to $\texttt{c}[\stackrel{\rightarrow}{i}] = ...$.

\begin{algorithmic}

\STATE \COMMENT {Begin of outer loop $i$.}
\STATE {$\texttt{index}_0[i] = \phi(\texttt{index}_1[i-1], -1)$} 
\STATE {$\texttt{min}_0[i] = \phi(\texttt{min}_1[i-1], 10000)$}

\STATE {$\texttt{C}'[i,j] = \mathit{raise\_dim}(\texttt{C}[j],i)$}
\STATE {$\texttt{d}[\stackrel{\rightarrow}{i},\stackrel{\rightarrow}{j}] = \mathit{SUM}(\texttt{S}[\stackrel{\rightarrow}{i},\stackrel{\rightarrow}{j}], \texttt{C}'[\stackrel{\rightarrow}{i},\stackrel{\rightarrow}{j}])$}
\STATE {$\texttt{p}[\stackrel{\rightarrow}{i},\stackrel{\rightarrow}{j}] = \mathit{MUL}(\texttt{d}[\stackrel{\rightarrow}{i},\stackrel{\rightarrow}{j}],\texttt{d}[\stackrel{\rightarrow}{i},\stackrel{\rightarrow}{j}])$}

\STATE \COMMENT {Begin of inner loop $j$. Loop is now much "shorter"!}
\STATE {$\texttt{sum}_0[i,j] = \phi(\texttt{sum}_1[i,j], [0,0,...])$}
\STATE {FOR j=0; j<=D; j++} \COMMENT{ Will turn into a MOTION loop.}
\STATE \hspace{0.25cm}{$\texttt{sum}_1[\stackrel{\rightarrow}{i},{j}] = \mathit{ADD}(\texttt{sum}_0[\stackrel{\rightarrow}{i},{j}],\texttt{p}[\stackrel{\rightarrow}{i},{j}])$}
\STATE \COMMENT {End of inner loop $j$.}

\STATE {$\texttt{sum}'_1[i] = \mathit{drop\_dim}(\texttt{sum}_1[i,j],j)$}
\STATE {$\texttt{c}[\stackrel{\rightarrow}{i}] = \mathit{CMP}(\texttt{sum}'_1[\stackrel{\rightarrow}{i}],\texttt{min}_0[\stackrel{\rightarrow}{i}])$}
\STATE {$\texttt{index}_1[\stackrel{\rightarrow}{i}] = \mathit{MUX}(\texttt{c}[\stackrel{\rightarrow}{i}],\texttt{index}_1[\stackrel{\rightarrow}{i}],[0,1,...N])$}
\STATE {$\texttt{min}_1[\stackrel{\rightarrow}{i}] = \mathit{MUX}(\texttt{c}[\stackrel{\rightarrow}{i}],\texttt{sum}'_1[\stackrel{\rightarrow}{i}],\texttt{min}_1[\stackrel{\rightarrow}{i}])$}
\STATE \COMMENT {End of outer loop $i$.}

\end{algorithmic}
